---
title: "Great Circles purrr"
author: "Jesse Sadler"
date: "2/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document shows different ways to create `sfc_LINESTRING` using `purrr` functions instead of a for loop. There are a couple of different ways. One of the two main ways that I have found uses both `base` and `purrr` functions. The other uses only `purrr`. I will first set up the script and then show the for loop for comparison.

```{r set-up, message = FALSE}
library(tidyverse)
library(sf)

letters <- read_csv("data/dvdm-correspondence-1591.csv")
locations <- read_csv("data/locations-1591.csv") %>% 
  select(place:lat) # simplify locations data to only necessary variables

routes <- letters %>%
  group_by(source, destination) %>% 
  summarise() %>% 
  drop_na() %>% # remove rows in which source or destination is unknown
  ungroup()
```

The method that uses `base` and `purrr` uses an input of latitude and longitude values in a matrix. The `purrr` method uses an input in the form of a data frame. Therefore, I need to make both sources and destinations matrices and data frames.

```{r source and destination data}
# data frames
sources_tbl <- routes %>% 
  left_join(locations, by = c("source" = "place")) %>% 
  select(lon:lat)
destinations_tbl <- routes %>% 
  left_join(locations, by = c("destination" = "place")) %>% 
  select(lon:lat)

# matrices
sources_m <- routes %>% 
  left_join(locations, by = c("source" = "place")) %>% 
  select(lon:lat) %>% 
  as.matrix()
destinations_m <- routes %>% 
  left_join(locations, by = c("destination" = "place")) %>% 
  select(lon:lat) %>% 
  as.matrix()
```

## For loop
The for loop method is probably the most straightforward in the it transparently vectorizes creating a LINESTRING, creating a list of LINESTRING sfg objects, which can then be turned into an sfc_LINESTRING object with `st_sfc()`.

```{r for loop linestring}
linestrings_sfg <- vector("list", nrow(routes))

for (i in 1:nrow(routes)) {
  linestrings_sfg[[i]] <- st_linestring(rbind(sources_m[i, ], destinations_m[i, ]))
}

linestrings_sfc <- st_sfc(linestrings_sfg, crs = 4326)
```

## Method 1: Separate source and destinations
The first method uses `base::split` to split the rows of the source and destination matrices into lists, which can then be put into the correct format of a 2x2 matrix with `purrr::map`. This can be done with sources and destinations as separate matrices or column binded into a single matrix. First, I will show it done as separate matrices, which necessitates the use of `purrr::map2()`.

```{r method 1}
# split matrices to create list of length equal to nrow sources_m
sources_list <- split(sources_m, seq(nrow(sources_m)))
destinations_list <- split(destinations_m, seq(nrow(destinations_m)))

# Bind rows of the lists
map2_list <- map2(sources_list, destinations_list, rbind)
# Need to use map to apply st_linestring on each item of the list
map2_sfg <- map(map2_list, st_linestring)

# sfc object
map2_sfc <- st_sfc(map2_sfg, crs = 4326)
```

## Method 2: column bound sources and destinations
This method is the same as method 1, but it begins by column binding the sources and destinations matrices, so that you do not need to use `map2`.

```{r method 2}
# Bind the columns
bound_m <- cbind(sources_m, destinations_m)

# Split into list of 1x4 matrix
bound_split <- split(bound_m, seq(nrow(bound_m)))
# Modify 1x4 matrix to 2x2, byrow gives columns of lon and lat
bound_list <- map(bound_split, ~ matrix(., nrow = 2, byrow = TRUE))
# Same need to use map to apply st_linestring on each item of the list
bound_sfg <- map(bound_list, st_linestring)

# sfc object
bound_sfc <- st_sfc(bound_sfg, crs = 4326)
```

## Method 3: pure `purrr`
The third method uses all `purrr` functions to do the same basic workflow as above. In the place of `split()` this method uses `transpose()`, which takes a data frame and creates a list of each row with a list of each column. Because `transpose()` creates a list of lists instead of a list of matrices, you need to `flatten()` on the second level list and create a properly formatted matrix, both of which can be done within a `map()` function. This method needs a data frame as an input, because `transpose()` does not work with matrices.

The only difference between this method and the first two as far as output is that the first two methods name the lists with integers, whereas `transpose()` does not name the vectors.

```{r method 3}
# bind columns in sources and destinations tibbles
bound_tbl <- bind_cols(sources_tbl, destinations_tbl)

# Turn bound_tbl into a list of rows with a list of columns
trans <- transpose(bound_tbl)

# Flatten second-level list into vectors and create matrices
trans_list <- map(trans, ~ matrix(flatten_dbl(.), nrow = 2, byrow = TRUE))

# Same need to use map to apply st_linestring on each item of the list
trans_sfg <- map(trans_list, st_linestring)

# sfc object
trans_sfc <- st_sfc(trans_sfg, crs = 4326)
```

Methods 2 and 3 are pipeable

```{r}
method2_sfc <- bound_m %>% 
  split(seq(nrow(bound_m))) %>% 
  map(~ matrix(., nrow = 2, byrow = TRUE)) %>% 
  map(st_linestring) %>% 
  st_sfc(crs = 4326)

method3_sfc <- bound_tbl %>% 
  transpose() %>% 
  map(~ matrix(flatten_dbl(.), nrow = 2, byrow = TRUE)) %>% 
  map(st_linestring) %>% 
  st_sfc(crs = 4326)
```

